---
title: "DATA 607 Extra Credit #1"
author: "Kristin Lussi"
date: "9/3/2023"
output: html
editor: visual
---

## What are the most common symptoms of people with lung cancer?

```{r}
library(readr)
library(gt)
library(dplyr) 
library(quarto)
library(ggplot2)

urlfile="https://raw.githubusercontent.com/kristinlussi/607ExtraCredit1/main/survey%20lung%20cancer.csv"

lungcancer<-read_csv(url(urlfile), show_col_types = FALSE)
```

Here is an overview of the data:

```{r}
head(lungcancer)
```

Here we will create subset of the data containing only the individuals who have lung cancer.

```{r}
cancer <- subset(lungcancer, LUNG_CANCER == "YES")
head(cancer)
```

The symptoms that are recorded in this data set are: yellow fingers, fatigue, wheezing, coughing, shortness of breath, difficulty swallowing, and chest pain.

Here, we will calculate the percentage of individuals who suffer from each symptom.

```{r}
# find number of rows
rownum <- nrow(cancer)

# find number of individuals who have yellow fingers
yellowfingers1 <- length(which(cancer$YELLOW_FINGERS==2))
# find percentage of individuals who have yellow fingers
yellowfingers <- yellowfingers1 / rownum

# find number of individuals who have fatigue
fatigue1 <- length(which(cancer$FATIGUE==2))
# find percentage of individuals who have fatigue
fatigue <- fatigue1 / rownum

# find number of individuals who have wheezing
wheezing1 <- length(which(cancer$WHEEZING==2))
# find percentage of individuals who have wheezing
wheezing <- wheezing1 / rownum

# find number of individuals who have coughing
coughing1 <- length(which(cancer$COUGHING==2))
# find percentage of individuals who have coughing
coughing <- coughing1 / rownum

# find number of individuals who have shortness of breath
shortbreath1 <- length(which(cancer$`SHORTNESS OF BREATH`==2))
# find percentage of individuals who have shortness of breath
shortbreath <- shortbreath1 / rownum

# find number of individuals who have difficulty swallowing
swallow1 <- length(which(cancer$`SWALLOWING DIFFICULTY`==2))
# find percentage of indiviudals who have difficulty swallowing
swallow <- swallow1 / rownum

# find number of individuals who have chest pain
chestpain1 <- length(which(cancer$`CHEST PAIN`==2))
# find percentage of individuals who have chest pain
chestpain <- chestpain1 / rownum 
```

Here we will create a table showing the different symptoms and their corresponding percentages.

```{r}
symptoms <- c("Yellow Fingers", "Fatigue", "Wheezing", "Coughing", "Shortness of Breath", "Difficulty Swallowing", "Chest Pain")

percentages <- c(yellowfingers, fatigue, wheezing, coughing, shortbreath, swallow, chestpain)

symptoms_df <- data.frame(symptoms, percentages)

symptoms_df[, 2] <- round(symptoms_df[, 2], digits = 4) * 100

symptoms_df <- symptoms_df[order(-symptoms_df[, 2]), ]

symptoms_tbl <- gt(symptoms_df)

symptoms_tbl <- 
  symptoms_tbl |>
  tab_header(
    title = md("**Percentage of Individuals with Lung Cancer who have Certain Symptoms**")
  ) |>
  cols_label(
    symptoms = md("**Symptoms**"), percentages = md("**Percentage of Individuals (%)**")
    )

symptoms_tbl
```

As you can see from the above table, the most common symptom that people with lung cancer experience is fatigue, with about 70% of individuals experiencing this symptom. The next most common symptom is shortness of breath, with 65.19% of individuals experiencing this symptom. The least common symptom that people with lung cancer experience is difficulty swallowing, with 51.85% of individuals experiencing this symptom. In conclusion, the top 3 most common symptoms for people with lung cancer are fatigue, shortness of breath, and coughing.

Just out of curiosity, let's visualize the symptom occurrence in individuals with lung cancer vs. the symptom occurrence in individuals without lung cancer.

```{r}
nocancer <- subset(lungcancer, LUNG_CANCER == "NO")

# find number of rows
rownum <- nrow(nocancer)

# find number of individuals who have yellow fingers
yellowfingers2 <- length(which(nocancer$YELLOW_FINGERS==2))

# find number of individuals who have fatigue
fatigue2 <- length(which(nocancer$FATIGUE==2))

# find number of individuals who have wheezing
wheezing2 <- length(which(nocancer$WHEEZING==2))

# find number of individuals who have coughing
coughing2 <- length(which(nocancer$COUGHING==2))

# find number of individuals who have shortness of breath
shortbreath2 <- length(which(nocancer$`SHORTNESS OF BREATH`==2))

# find number of individuals who have difficulty swallowing
swallow2 <- length(which(nocancer$`SWALLOWING DIFFICULTY`==2))

# find number of individuals who have chest pain
chestpain2 <- length(which(nocancer$`CHEST PAIN`==2))

# create the data frame
symptoms <- c("Yellow Fingers", "Fatigue", "Wheezing", "Coughing", "Shortness of Breath", "Difficulty Swallowing", "Chest Pain")

cancercount <- c(yellowfingers1, fatigue1, wheezing1, coughing1, shortbreath1, swallow1, chestpain1)

nocancercount <- c(yellowfingers2, fatigue2, wheezing2, coughing2, shortbreath2, swallow2, chestpain2)

count_df <- data.frame(symptoms, cancercount, nocancercount)

# Reshape the data for plotting
library(reshape2)
count_df <- melt(count_df, id.vars = "symptoms")

# Create a bar plot
ggplot(count_df, aes(x = symptoms, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Symptom Counts for Individuals with Lung Cancer vs. No Lung Cancer",
       x = "Symptom",
       y = "Count of Individuals") +
  scale_fill_manual(values = c("Lung Cancer" = "red", "No Lung Cancer" = "blue")) +
  theme_minimal()

```

As you can see, these symptoms are far more common in people with lung cancer. From this table, we can infer that these symptoms are likely caused by lung cancer.
